---
title: "ggplot2"
author: "Kaila Velasco"
date: '2023-10-05'
output: html_document
---

# Hands-on Outputs

## Output 1

**Using the same dataset, create a histogram for the variable area, and set the bins to 10.**

```{r}
library(ggplot2)
ggplot(data = rock, aes(x = area)) +
    geom_histogram(bins = 10) + 
    labs(x = 'Area')
```

## Output 2

**Which variable was used as the basis of the grouping?** Clarity

# Exercises

## Exercise 1
**Using the *mtcars* dataset, is there a significant difference in fuel efficiency based on transmission type?**

```{r}
ggplot(mtcars, aes(y = mpg, group = am, fill = am)) +
    geom_boxplot() + labs(title = 'Miles per gallon',
                          x = 'Transmission Type', 
                          y = 'Fuel efficiency')
```

## Exercise 2
**Using the *InsectSprays* dataset, which brand appears to be the most effective?**

```{r}
ggplot(data = InsectSprays, aes(y = count, x = spray, fill = spray)) +
	geom_bar(stat = 'summary', fun = 'sum') +
	labs(title = 'Insect Kill Count of Each Spray Brand')
```

# Exercise 3

**Using the *trees* dataset, which of the two variables exhibit a linear relationship?**

```{r}
ggplot(data = trees,
       aes(x = Girth, y = Volume)) + 
    geom_point() + 
    labs(x = "Tree diameter", y = "Volume of tree",
         title = "Relationship between tree diameter and volume") +
    geom_smooth(se = F)
```

